---
title: "How to keep it togetheR"
author: "Marc A.T. Teunis, Ph.D."
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    highlight: pygments
    widescreen: yes
subtitle: List-columns in a dataframe
resource_files:
- data/CISID_pertussis_10082018.csv
- images/open_compaq.jpg
- images/pepper.jpg
- R/do_tidy_pertussis.R
- resources.html
- resources.Rmd
- README.md
---

<style>

.footer {
    font-size: 16pt;
    color: Black;
    font-family: 'Calibri'
    content: attr(label);
    position: absolute;
    margin-left: auto;
    margin-right: auto;
    left: 0;
    right: 0;
    text-align:center;
    bottom: 10px;
    line-height: 1.9;
    display: block;
}

</style>

<div class="footer" style="margin-top;font-size:120%;"> 
Using list-columns - R Cafe UU, 28 January 2019 </div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE, 
                      error = FALSE, 
                      fig.width = 4,
                      fig.height = 4)
```

```{r, root, include=FALSE, eval=FALSE}
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)

```

## Who am I?

 - Background in (Medical) Biology
 - PhD in (Psycho)Neuroimmunology (WKZ - UMCU)
 - Validation and development of in-vitro methods for toxicology
 - **SpaRked my interest in R**

## What is R to me?

 - Every day go-to for analytics of very different types of data
 - Statistical analysis (validation / in vitro studies)
 - Genomics (`DESeq2`, `edgeR`, `limma`, `Glimma` )
 - Microbiome analysis (`Bioconductor` & `qiime2`)
 - Interactivity with `Shiny` and `flexdashboards`
 - Reproducibility (`rmarkdown`, )


## Contents

List-columns and managing many models with `{tidyverse}`

 1. Recap: Dataframes, Lists and Iterations
 2. `map()` functions from the `{purrr}` package
 3. Using list-colums (in a dataframe)
 4. Implementing list-columns to create models and graphs

 PART I: FIRST STEPS (for today) 
 
 PART II: MORE LIST-COLUMNS (do-it-yourself) 

# PART I: FIRST STEPS (slide 7-)

Clone the repository from:

`https://github.com/uashogeschoolutrecht/rcafe` or login

`http://rcafe.hudatascience.nl` with user credentials on little piece of paper

## Packages
The packages used in this tutorial
```{r, packages}
library(tidyverse)
library(modelr)
library(lubridate)
library(broom)
library(purrr)
library(repurrrsive)
```

## Dataframes and Lists are recursive vectors
```{r, recursive_vectors}
table <- tribble(
  ~ a, ~b, ~c, ~d, 
  "x", 1, TRUE, 1.45, 
  "y", 2, FALSE, 3.88,
  "z", 3, TRUE, 33.5
  ) 
table
```

## Checking characteristics of `table`
```{r}
is.vector(table)
is.vector(table$a)
is.atomic(table)
is.atomic(table$c)
is.recursive(table)
class(table)
typeof(table)
## So dataframes are a special kind of lists?!? 
```

## Column containing a list (in a dataframe)
```{r}
table <- tribble(
  ~ a,      ~b,      ~c,       ~d,     ~e, 
  "x",      1,       TRUE,     1.45,   numbers = 1:10,
  "y",      2,       FALSE,    3.88,   lgl = c(TRUE, FALSE),
  "z",      3,       TRUE,     33.5,   city = "Utrecht"   
  ) 
table
```

## Iterate over a dataframe
```{r}
map(table, class)
```

## Iterate over a list
```{r}
map(table$e, nchar)
```

# Case data

Let's switch to RStudio and open the file: 

`demo.Rmd`

## Data origin

Whooping cough outbreaks from The World Health Organization

http://data.euro.who.int/cisid/?TabID=463987

See for variable and data collection also: 

http://ecdc.europa.eu/sites/portal/files/documents/Pertussis%20AER.pdf

for more details see file: "load_data.R"

## Load the tidy version of the dataset
The code for cleaning and tidying the data is in the file "./load_data.R"
```{r, read_data}
source(file = file.path("load_data.R"))
  
```

## Inspect data
```{r, inspect_data}
head(pertussis_data_tidy, n = 2)
names(pertussis_data_tidy)
```

## For which years is there data recorded?
```{r, echo=FALSE, fig.width=4, fig.height=4}
pertussis_data_tidy %>%
  ggplot(aes(x = year, y = annual_pertussis_cases)) +
  geom_point(aes(position = "jitter" ))
```

## Log10 cases by year, `colour = country`
```{r, echo=FALSE, fig.width=4, fig.height=4}
pertussis_data_tidy %>%
  na.omit() %>%
  group_by(country) %>%
  ggplot(aes(x = year,
             y = log10(annual_pertussis_cases))) +
  geom_line(aes(group = country, colour = country))  +
  theme_bw() +
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
This is a messy graph indeed!!

## Only the Netherlands
```{r, echo=FALSE, fig.width=4, fig.height=4}
## only Netherlands data
netherlands_data <- pertussis_data_tidy %>%
  dplyr::filter(country == "Netherlands") %>%
  na.omit() ## for sake of simplicity we remove the years that are missing

netherlands_data %>%
  ggplot(aes(x = year,
             y = annual_pertussis_cases)) + 
  geom_line() +
  geom_point(colour = "red") +
  ggtitle("The Netherlands; Whooping Cough cases")
```

## Plot linear model for NL
```{r, echo=FALSE, fig.width=4, fig.height=4}
netherlands_model <- lm(
  annual_pertussis_cases ~ year, 
  data = netherlands_data %>%
    na.omit()
)
## plot model for NL

netherlands_data %>%
  na.omit() %>%
  add_predictions(netherlands_model) %>%
  ggplot(aes(x = year, 
             y = pred)) +
  geom_line() +
  ggtitle("Linear trend") +
  geom_point(
    data = netherlands_data,
    aes(x = year, 
    y = annual_pertussis_cases),
    colour = "red") +
  geom_line(
    data = netherlands_data %>% na.omit, ## note the pipe to remove NA
    aes(x = year, 
        y = annual_pertussis_cases),
    colour = "red",
    linetype = "dashed"
)

```

## How can we apply this to `every` country 
without doing the above `r pertussis_data_tidy$country %>% as_factor() %>% levels() %>% length()` times.

## Split the data by country and apply the model or graphing code to each subset

 - In fact, data for each country is a subset of the full dataset
 - We can subset the original dataframe into seperate dataframes for each country 

## Using list-columns to keep track of your results and models
```{r}
nested_pertussis <- pertussis_data_tidy %>%
  na.omit() %>%
  dplyr::select(country, year, annual_pertussis_cases) %>%
  group_by(country) %>%
    nest() 

head(nested_pertussis, 2) ## you see the grouping by country
head(nested_pertussis$data[[1]], 2) ## you get the individual country df
```

## Label (name) the idividual elements of the list column 
```{r, eval=FALSE} 
names(nested_pertussis$data) <- nested_pertussis$country 
```
 
## Linear model for each country
First we write a function that creates the linear model for  one country
```{r}
country_model_lm <- function(df){
  model <- lm(
    annual_pertussis_cases ~ year, 
    data = df)
  return(model)
}
```

## Iterate the model function over nested `$data` with `purrr::map()`
```{r}
models <- map(
  nested_pertussis$data, country_model_lm
  )
head(models, 2)
```

## Extracting information from models the tidy way
```{r}
# models[[1]] %>% summary 
models[[1]] %>% broom::glance()
models[[1]] %>% broom::glance() %>%
  pluck("adj.r.squared")
```

## Keep it togetheR
 - We have the models now
 - Better to store them together with the data
 - By using `dplyr::mutate()` 
 
## Create an additional list-column on the basis of an existing one
```{r}
nested_pertussis <- nested_pertussis %>%
  mutate(models_lm = map(data, country_model_lm))
# View(nested_pertussis)
head(nested_pertussis, 2)
```

## Add model summaries as a list column
```{r}
nested_pertussis <- nested_pertussis %>%
  mutate(models_lm_summary = map(models_lm, summary))
# View(nested_pertussis)
```

## Extracting information
```{r}
nested_pertussis <- nested_pertussis %>%
  mutate(params_lm =  map(models_lm, broom::glance)) %>%
  mutate(p_value = map(params_lm, pluck, "p.value"))

nested_pertussis$params_lm[[1]]
nested_pertussis$p_value[[1]] %>% round(6)

```

## Adding residuals to all the models
```{r}
nested_pertussis <- nested_pertussis %>% 
  mutate(
    resids = map2(data, models_lm, add_residuals)
  )
```

## `map2()`

 - What type of arguments does the function `map2()` take?
 - How would you find more information?

## Unnest the residuals
```{r}
resids <- nested_pertussis %>%
  unnest(resids)
head(resids, 2)
```

## Plotting all residuals for each country
```{r}
plot_resids <- function(resids){
  
  plot <- resids %>% 
  ggplot(aes(year, resid)) +
    geom_line() + 
    geom_smooth(se = FALSE) +
    theme_bw() +
    theme(legend.position="none") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

  return(plot)
}


```

## Nest all plots inside a new list column
```{r, fig.width=4, fig.height=4}
nested_pertussis <- nested_pertussis %>%
  mutate(resids_plot = map(resids, plot_resids))

## name alements in the list-column
names(nested_pertussis$resids_plot) <- nested_pertussis$country

## show one plot
nested_pertussis$resids_plot$Belarus
```

## What can we conclude from these graphs?
```{r, fig.width=3, fig.height=3}
nested_pertussis$resids_plot$Belgium
nested_pertussis$resids_plot$Armenia
```

## Learn more?

'Managing many models with R' by Hadley Wickham - Lecture
https://www.youtube.com/watch?v=rz3_FDVt9eg

'R for Data Science' by Garret Grolemund & Hadley Wickham
https://r4ds.had.co.nz/
Especially chapters:
21 - https://r4ds.had.co.nz/iteration.html
25 - https://r4ds.had.co.nz/many-models.html

Study part II below on your own...

##
```{r, dpi=40, echo=FALSE}
knitr::include_graphics(path = file.path("images",
                                         "pepper.jpg"))
```












# PART II; Extracting more information from a list column 

***ADVANCED***

For another day...

## Looking at quantative statistical measures for model quality
```{r}
r_squared <- nested_pertussis %>% 
  mutate(glance = map(models_lm, broom::glance)) %>% 
  unnest(glance, drop = TRUE) %>%
  select(country, r.squared, AIC, BIC) %>%
  arrange(r.squared)
```

## Plotting r.sqared values
```{r}
r_squared %>%
  ggplot(aes(x = reorder(as_factor(country), r.squared),
             y = r.squared)) +
  geom_point() +
  coord_flip() +
  ylab("R-squared") +
  xlab("Countries") + 
  geom_point(data = filter(r_squared, r.squared <= 0.25), colour = "red")
```

## Plotting pertussis cases for countries with low r.squared over time
 1. Filter countries with `r.squared` <= 0.25
 2. Put  countries in vector
 3. Plot data

## Step 1
```{r}
low_r_squared <- r_squared %>%
  dplyr::filter(r.squared <= 0.25) %>%
  dplyr::select(country) 
low_r_squared <- low_r_squared$country
```

## Step 2
```{r}
low_r_squared_nested <- nested_pertussis %>%
  dplyr::filter(country %in% low_r_squared) %>%
  select(country, data) %>%
  unnest()
```

## Step 3
```{r}
low_r_squared_nested %>%
  ggplot(aes(x =year, y = annual_pertussis_cases)) +
  geom_line(aes(group = country, color = country)) +
    theme_bw() +
  theme(legend.position="none") +
  facet_wrap( ~ country, scales = "free") +
 theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
What is happening with the pertussis vaccination grade over the past 8 years? 

## Store ggplot2 objects in a list-column
 1. Create a function that makes the plot
 3. Test function on single dataframe
 2. Apply the function using `mutate()` and `map()` to all dataframes or models
 
```{r}
## isolate one dataframe to test function
df <- nested_pertussis$data[[1]]
plot_country <- function(df){
  
  df %>%
    ggplot(aes(x = year,
           y = annual_pertussis_cases)) +
    geom_line() +
    geom_smooth() +
    ylab("Annual cases")

}

## test function
plot_country(df = df)
```

## Apply plotting function to nested data
```{r}
nested_pertussis <- nested_pertussis %>%
  mutate(plots_cases_over_time = map(data, 
                                     plot_country))

```

## Add countries as names to nest-column
```{r}
names(nested_pertussis$plots_cases_over_time) <- 
  c(nested_pertussis$country)

nested_pertussis$plots_cases_over_time[1]
```

## Pull out "The Netherlands"
```{r}
nested_pertussis$plots_cases_over_time$Netherlands
```

## Plotting a panel of 4 graphs
```{r, echo=FALSE}
grid <- cowplot::plot_grid(
  nested_pertussis$plots_cases_over_time$Armenia, 
  nested_pertussis$plots_cases_over_time$Austria,
  nested_pertussis$plots_cases_over_time$Ukraine,
  nested_pertussis$plots_cases_over_time$Netherlands, 
  labels = c("Armenia", "Austria", "Ukraine", "Netherlands"),
  scale = 0.7, hjust = c(-0.7, -0.8, -0.8, -0.5))
grid
```

## Literature
For some background on the pattern we are seeing

 - https://www.scientificamerican.com/article/why-whooping-cough-vaccines-are-wearing-off/
 - http://outbreaknewstoday.com/pertussis-cases-up-significantly-in-the-eu-netherlands-and-uk-worst-hit-55315/


## Exploring many more models
Let's add a quadratic model in the mix. Assume we want to explore non-linear relationships in this dataset
```{r}
non_linear_model <- function(df, model_params){
  
  nl_model <- lm(
    annual_pertussis_cases ~ poly(year, 
                                  model_params),
                 data = df)
  
  return(nl_model)
}
```

## Creating a safe version of this function
```{r}
safe_non_linear <- purrr::safely(non_linear_model)
## apply test:
df = nested_pertussis$data[[1]]

test_non_linear <-  df %>% non_linear_model(df = .,
                                            model_params = 2)
```

## Test function on one country
```{r}
test_non_linear %>% broom::glance()
```

## Apply model to all countries 
We rerun the steps above to add this new model and new graphs to the nested dataframe

Add new model to the nested table
```{r}
nested_pertussis <- nested_pertussis %>%
  mutate(models_nl_2 = map(data, safe_non_linear, 
                          model_params = 2))
  
nested_pertussis$models_nl_2 <- transpose(nested_pertussis$models_nl_2)
```

## Set names to elements in the list-column
To be able to `pluck()` by name later
```{r}
names(nested_pertussis$models_nl_2$result) <- nested_pertussis$country
```

## Pluck results in new list-column
```{r}
nested_pertussis$models_nl_2$result

nested_pertussis <- nested_pertussis %>%
  mutate(statistics_nl = pluck(models_nl_2, "result"))

nested_pertussis$statistics_nl[[1]]

```

## Tidy the list-column with `{broom}`
```{r}
nested_pertussis <- nested_pertussis %>%
  mutate(parameters_nl = map(statistics_nl, glance))
```


## Looking at quantative statistical measures for model quality
```{r}
r_squared_nl <- nested_pertussis %>%
  select(country, parameters_nl) %>%
  unnest()
```

## Plotting r.sqared values
```{r, echo=FALSE}
r_squared %>%
  ggplot(aes(
    x = reorder(as_factor(country), r.squared),
    y = r.squared
  )) +
  geom_point() +
  ylab("Countries") +
  geom_point(data = filter(
    r_squared,
    r.squared <= 0.25
  ), colour = "red") +
  geom_point(
    data = r_squared_nl,
    aes(
      x = reorder(as_factor(country), r.squared),
      y = r.squared
    ), colour = "purple",
    shape = 21, size = 3
  ) +
  theme_bw() +
  xlab("Country") +
  ylab("R.squared") +
  coord_flip() +
  geom_point(data = r_squared_nl %>%
               dplyr::filter(country == "Ireland" |
                             country == "Belgium"), 
             colour = "blue", shape = 22, size = 5)



```

## Let's examine two models for two countries where the non-linear did and did not not improve the R.squared: Ireland (improved) and Belgium (not-improved)  
```{r}

x <- nested_pertussis %>%
  select(country,
         data, 
         models_lm,
         statistics_nl) %>%
  gather(models_lm:statistics_nl, key = "models", value = "model_params") %>%
 print()  
```

## Remove 'empty model'
```{r}
ind <- x$model_params == "NULL"
#ind <- x$data == "NULL"
x <- x[!ind, ]
```

## Add prediction-list column
```{r}
predictions <- x %>%
#  filter(country == "Czech Republic") %>%
  mutate(predictions = map2(data, model_params, add_predictions,
         var = "annual_pertussis_cases")) %>%
  filter(country == "Ireland" | 
         country == "Belgium") %>%
  select(country, data, predictions)

```

## Set names
```{r}

names(predictions$predictions) <- predictions$country
```

## Belgium
```{r}
nested_pertussis$plots_cases_over_time$Belgium +
  geom_line(data = predictions$predictions[[1]], colour = "red", size = 2) +
  
  geom_line(data = predictions$predictions[[3]], colour = "green", size = 2) 
```  

## Ireland
```{r}
nested_pertussis$plots_cases_over_time$Ireland+
  geom_line(data = predictions$predictions[[2]], colour = "red", size = 2) +
  
  geom_line(data = predictions$predictions[[4]], colour = "green", size = 2) 
```


## Disclaimer & Licence

_The work in this presentation has been adapted from many sources as mentioned in slide 'Credits'. For a full reference list see the 'REFERENCES.txt' document in the github repo 
https://www.github.com/usashogeschoolutrecht/surf_research_bootcamp_2018

The work presented here may be shared, remixed or adapted as long as the original references and the authors of this document are mentioned in the redistribution: CC BY-SA


## Credits

Much of this material has been derived in one way or the other from the work of Hadley Wickham and Garret Grolemund and many others. For a more elaborate reference list see the REFERENCES.txt file in the project root.

Thanks to Hadley & Garret for writing the book "R for Data Science" http://r4ds.had.co.nz/ and for their work in general to innovate the R world.

Work on integration of Git/Github with R/RStudio is thoroughly and wit-fully documented by Jenny Brian. I also very much appreciate her work on the use of RMarkdown. See also:

https://github.com/jennybc/happy-git-with-r & http://stat545.com/block007_first-use-rmarkdown.html 
