---
title: "Start with RMarkdown"
subtitle: "A gentle slide (show) into R-package development"
author: "Marc A.T. Teunis"
date: "`r Sys.time()`"
always_allow_html: true
output: 
  beamer_presentation:
    theme: "AnnArbor"
    colortheme: "dolphin"
    fonttheme: "structurebold"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE)
```

## Contents

 1. Writing functions
 1. "Start with Rmd" - What is it?
 1. Function documentation - `{roxygen2}`
 1. Demo - Building a package `{usethis}`, `{devtools}`
 
## Prerequisites

 - Windows: Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/)
 - Clone Materials from:
 https://github.com/UtrechtUniversity/R-data-cafe/start_with_rmd
 - Install packages:

```{r}
library(tidyverse)
library(here)
library(usethis)
library(devtools)
library(reprex)
```

## Intro

 **Discuss with your neighbour (2 min.)**
 
 1. When was the last time you wrote a function in R? 
 1. What did it do?

## R Markdown

 - An 'R' implementation of the simple Markdown mark-up language 
 - Combines prose with R code and output
 - Many output formats
 - Accepts html, css, LaTeX
 - Is generally a good starting point when setting up an analysis
 - As it turns out: it is also a good starting point for creating an R-package

## Why start with RMarkdown (for an R-package)

 1. RMarkdown enforces more elaborate documentation (narratives + code + output)
 1. Analysis follows a natural flow
 1. Functions are put into context of application
 1. Information, analysis and functions are collected in one place
 1. When creating a package, you already have a vignette (documentation) 

## Why build an R-package

 1. R-package is the natural 'communication' vessel for R-code
 1. Documentation increases reproducibility (future you will be grateful)
 1. R-packages are the natural habitat for R functions
 1. Do not repeat yourself (code + data + documentation all in one place without redundancy)
 
## Imagine this data

```{r}
data_dengue <- read_csv(
  here::here(
    "data-raw", "dengue_data.csv"), 
  comment = "#")

data_dengue[c(1:5), c(7:11)]  
```

## Discuss with your neighbour!
What does this part do?
```
here::here(
    "data-raw",
    "D010",
    "dengue_data.csv"
  )
```

## Why is it handy?

see ["here! here!" - github link J. Bryan](https://github.com/jennybc/here_here)

__Please do no do this in your code, anywhere:__ 

```
data_dengue <- read_csv(
  "D:/r_projects/start_with_rmd/data-raw/dengue_data.csv", 
  comment = "#")
```

Use the `{here}` package instead

## What is typical about the missing values?

```{r}
data_dengue[c(1:5), c(7:11)] 
```

_[Reference: Hadley Wickham - link](https://richmedia.lse.ac.uk/methodologyinstitute/20170814_ExpressYourselfUsingR.mp4)_

## We could solve it like this
```{r, eval=FALSE}
data_dengue$Venezuela[
  data_dengue$Venezuela ==  999] <- NA
data_dengue$Venezuela[
  data_dengue$Argentina ==  "missing"] <- NA
data_dengue$Philippines[
  data_dengue$Philippines == 999] <- NA
data_dengue$Philippines[
  data_dengue$Philippines == 990] <- NA
```

## Discuss with your neighbour

Why is this syntax a bad idea?
```
data_dengue$Philippines[
  data_dengue$Philippines == 990] <- NA
```

## DRY - Don't Repeat Yourself

 1. When entering the 'same' code twice, it is time to start writing a function
 1. Copying & pasting is error prone
 1. Extend this to files, workflows and data-copies
 1. Keep everything that belongs together in the same place:
 
 **"Start with Rmd"**
 
 `RStudio Project --> RMarkdown file --> R-package`

## Functions in R

```
descriptive_function_name <- function(arguments, ...){
 if(some_condition_on_argument{ 
   message("message that condition meets argument(s)")
 } else {
   warning/stop("condition does NOT meet criteria ")
 }
 
 * some operations on the function arguments
 * some more calculations, 
 * maybe reshaping the object or looping or transforming
 * ... = arguments that can be passed to functions 
 used inside a function

 return(whatever_the_function_returns)
}
```

## Putting the DRY principle to practice
The function below replaces certain values (`na_string`) 
in a vector for `NA`

```{r}
replace_x_for_na <- function(x, na_string){
  x[x %in% na_string] <- NA # %in% is a special for match
  return(x)
}
```

## Apply this function to one column of our `data_dengue`

```{r}
data_dengue$Venezuela[1:4]
replace_x_for_na(x = data_dengue$Venezuela, 
                 na_string = c(99, 990:999, "missing", "mising")) %>%
  as.numeric() %>%
  head(4)
```

## Apply our function to whole `data_dengue` dataframe

```{r}
data_dengue_new <- data_dengue %>%
  purrr::map_df(
  replace_x_for_na, 
  na_string = c(99, 990:999, "mising", "missing"))
data_dengue_new[c(1:5), c(7:11)]
```


## Function Documentation
Let's assume we want to write a bit of documentation on how to use our `replace_x_for_na()` function

## Documentation for functions
__How it is done in R-packages__
Let's look at the documentation for the function `mean`
```{r, eval=FALSE}
?mean
```

## Writing you own documentation: Roxygen Comments
To construct this formal documentation structure we can use Roxygen Comments (`#'`)

```
#' @title Change values into formal NA
#' 
#' @param x A vector containing non-formal NA values 
#' that need to be replaced by formal NA
#' @param na_string A vector of values indicating 
#' which values need to be replaced by NA
#' @return A mutated vector, with NA as replacements
#' @example <a reprex goes here>
#' @export

<function definition>

```

## Steps to build an R-package (from Rmd)

 1. Create a package project `usethis::create_package(<pkg_name>)`
 1. Extract functions from your Rmd file into the ~/R folder
 1. Write/Check documentation
 1. Add other package infrastructure
 1. Add dependencies
 1. Build documentation and vignettes
 1. Write tests --> Test package
 1. Check package
 1. Build package (source/binary)
 1. Publish package on Github and/or CRAN/Bioconductor
 
## Step 1: create an R-package RStudio project
Many steps of building an R-package are supported by the `{usethis}` package

 1. This will create a backbone for an R-package called 'bumblebee' in the root and will start up the RStudio project in a separate session
 1. Here we will put this package inside an existing RStudio project, usually this is a bad idea (but now we can more easily monitor whats going on)
 
```{r, eval=FALSE}
usethis::create_package(path = "bumblebee")
```

## Populating the /bumblebee/R/ folder

**Execute the following steps inside the ~/bumblebee project**

 1. Create a backbone function .R script for the function
  `replace_x_for_na()`
 with `usethis::use_r("replace_x_for_na")`
 1. Copy the function definitions + added Roxygen documentation to this .R file

## Generate documentation for the functions and the package
Run the following commands inside ~/bumblebee at the Console
```{r, eval=FALSE}
## create an .R file
usethis::use_r("replace_x_for_na")
```

copy function and documentation to this R/replace_x_for_na.R file
```
#' @title Change values into formal NA
#' 
#' @param x A vector containing non-formal NA values 
#' that need to be replaced by formal NA
#' @param na_string A vector of values indicating 
#' which values need to be replaced by NA
#' @return A mutated vector, with NA as replacements
#' @example <a reprex goes here>
#' @export
replace_x_for_na <- function(x, na_string){
  x[x %in% na_string] <- NA # %in% is a special for match
  return(x)
}

```

## Add general package documentation
```{r, eval=FALSE}
usethis::use_package_doc()
## build documentation for functions
devtools::document()
```

## Add dependencies
```{r, eval=FALSE}
## add the pipe es dependency so you can use it in you function defenitions 
usethis::use_pipe()
## add ggplot2 as a dependency
usethis::use_package("ggplot2")
usethis::use_package("dplyr")
usethis::use_package("readr")
usethis::use_package("here")
## update documentation
devtools::document()

```

## Add a licence
Here we will use CC-BY as an example
```{r, eval=FALSE}
usethis::use_ccby_license("Your Name")
```

## Add a dataset to a package
If you want to add a dataset:

_The raw-data has to be put inside the ~/bumblebee/data-raw folder._ 

The script that generates the dataset also goes inside the data-raw folder.
```{r, eval=FALSE}
usethis::use_data_raw()

## put the raw dataset inside 'data-raw' 
## copy the code below to the DATASET.R file:
DATASET <- readr::read_csv(
  here::here(
    "data-raw", "dengue_data.csv"),
  comment = "#")

DATASET <- DATASET %>%
  purrr::map_df(
    bumblebee::replace_x_for_na,
    na_string = c(99, 990:999, "mising", "missing")) %>%
  tidyr::gather(Argentina:Venezuela,
                key = "country",
                value = "cases")
usethis::use_data(DATASET, overwrite = TRUE)

```

## Building the documentation
```{r, eval=FALSE}
devtools::document()
```

## Building the package

**Click 'Build' --> 'Install & Restart'**

## Add a vignette
To add full documentation:
```{r, eval=FALSE}
usethis::use_vignette("bumblebee-demo")
```
You can then copy chunks from the current Rmd to this vignette. Of course you can leave out the function definitions because these will be avaialable from the package now.

## Rendering the vignette
```{r, eval=FALSE}
devtools::build_vignettes(pkg = "../bumblebee")
## load all to view the vignette
devtools::load_all(".")
## view the vignette
browseVignettes(package = "bumblebee")
```

# ADVANCED

## Writing tests with `testthat()`
Tests are important because:

 1. Things break
 1. Versions change (R and Dependencies)
 1. Inputs change
 
## How to start with tests
Again: `{usethis}` makes it easy to start 
```{r, eval=FALSE}
usethis::use_test("replace_x_for_na")
## test for datset
## test example
vec <- c(999, 998, "missing")
result <- bumblebee::replace_x_for_na(vec, na_string = c(999, 998, "missing"))

testthat::test_that("replace_x_for_na works!", {
  testthat::expect_equal(result, c(NA_character_,
                                   NA_character_,
                                   NA_character_))
})
```

## Test coverage
To see what still needs to be done ;-)
```{r, eval=FALSE}
usethis::use_travis()
usethis::use_coverage()
covr::report()
```

## Building and Checking package
Click:

 - 'Build' --> 
 - 'More' --> 
 - 'Clean & Rebuild' -->
 - 'Test Package' -->
 - 'Check Package' -->
 - 'Build Source Package' (Linux / McOS / Windows)
 - 'Build Binary Package' (Windows only)

## Integration with Git/Github

Initialize an empty repo in Github -->

In Terminal execute:
```
echo "# bumblebee" >> README.md
git init
git add README.md
git commit -m "first commit"
git remote add origin git@github.com:<gihub_username>/bumblebee.git
git push -u origin master
```
Commit changes locally --> push changes to remote

