---
title: "README"
author: "Marc A.T. Teunis"
date: "12/4/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      error = FALSE,
                      eval = FALSE,
                      message = FALSE)
```

This tutorial is about creating R-packages. It adheres to the so-called "Start with Rmd" principle, where you gradually build your analysis in an RMarkdown file and from there develop an R-package.

The full tutorial is available as `start_with_rmd.pdf` and `start_with_rmd.Rmd` file inside the folder `./Rmd`

## Contents

 1. Writing functions
 1. "Start with Rmd" - What is it?
 1. Function documentation - `{roxygen2}`
 1. Demo - Building a package `{usethis}`, `{devtools}`
 
## Prerequisites

 - Windows: Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/)
 - Clone Materials from:
 https://github.com/UtrechtUniversity/R-data-cafe/start_with_rmd
 - Install packages:

```{r}
library(tidyverse)
library(here)
library(usethis)
library(devtools)
```


## Online references

 - https://emilyriederer.netlify.com/post/rmarkdown-driven-development/
 - https://github.com/jennybc/pkg-dev-tutorial
 - https://r4ds.had.co.nz/
 - http://r-pkgs.had.co.nz/
 
 ## Good functions

 * Do one thing good
 * Have no 'side-effects'
 * `class(input) = class(output)` (to make the function 'pipe-able')

## Function documentation: We could do
```{r}
## A function that replaces a string or number to a 
## formal NA (missing value)
## this function takes a R-vector object x as 
## argument and returns a mutated vector in which 
## the indicated string or number is replaced with a 
## formal missing value

# <function definition goes here>

```

BUT: Roxygen rules!

## 'Many' functions? 
 
 1. When you find yourself writing many functions in an Rmd file like this
 1. It is time to start building an R-packge.
 
 ## Common workflow
```{r, echo=FALSE}
DiagrammeR::grViz("
digraph rmarkdown {
  'Edit documentation' -> 'devtools::document()'
  'devtools::document()' -> 'Adapt package structure'
  'Adapt package structure' -> 'Clean & Rebuild'
  'Clean & Rebuild' -> 'Check Tests'
  'Check Tests' -> 'Git Commit'
}
")
```